### 2013-11-12 src2/hl/paperThesis.tex
update

The NVIDIA CUDA capable GPUs are built on a scalable architecture. A GPU 
consists of a number of {\em multiprocessors}; each multiprocessor has a 
number of processing elements (cores) and a local memory that is shared 
between threads running on the cores. Figure \ref{fig:GPU} illustrates these 
concepts. A GPU can come with as few as one of these multiprocessors.
The two GPUs used in our measurements are the GTX 680, which has eight 
multiprocessors, with a total of 1536 processing cores, and the GT 650M, a 
laptop GPU, which has 384 such cores. On these cores, groups of 32 threads 
called {\em warps} are scheduled. There are a number of warp scheduling units 
per multiprocessor. Within a warp, threads execute in lockstep (SIMD); 
diverging branches, that is those that take different paths on different 
threads within a warp, are serialised, leading to performance penalties.

@@

Obsidian, like CUDA, differentiates between {\em Thread}, {\em Block} and 
{\em Grid} computations. The programmer specifies a sequential computation 
to be carried out by each thread. Many instances of a sequential computation,  
run in parallel over threads, to form a block. A number of such blocks run in 
parallel, to form a grid.

@@

This program does not make use of any specific details of the BProgram monad; 
instead it just composes {\tt return} and {\tt push} with the Obsidian library 
function {\tt reverse}. The {\tt push} function converts a pull array to a push 
array.

@@

\noindent
The {\tt pConcatMap} combinator is overloaded; here it is applied 
at the following type: 
%takes a local computation that results in a push array.
\begin{small}
\begin{Verbatim}[samepage=true] 
pConcatMap :: ASize l 
            => (a1 -> Program t (SPush t a))
            -> Pull l a1 
            -> Push (Step t) l a
\end{Verbatim}
\end{small}

@@

\noindent
where {\tt a0} to {\tt aN} can be pull arrays or scalars. This choice 
limits the kinds of kernels can be expressed, but
makes compilation of those kernels a lot easier.
For example, a kernel that outputs data to two different global
arrays cannot be expressed.

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\citehl{COMPILEEDSL} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\citehl{FELDSPAR} 
and Repa ~\citehl{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as needed from those available. 
If the array is larger than the number of threads made available, the 
force operation is split into two or more passes. This is a change compared
to earlier versions of Obsidian where the length of arrays we could force 
where limited to the maximum number of threads. In this version, the 
threads are, in a sense, virtual. The mapping from virtual threads to 
real threads is done as a step in the compilation.

@@

There are type synonyms for the dynamic and static versions of  pull arrays:

@@

The programs that can be specified in Obsidian
are represented in a {\tt Program} data type.
This data type also represents the lowest level programming interface 
exposed by Obsidian. In reference ~\citehl{JOSEFJOELCSORT} we make use of these low-level 
capabilities to implement histograms using atomic operations (not shown here).

@@

Below are some selected parts of the  {\tt Program} data type. There 
is a sequential, {\tt SeqFor}, loop that runs a sequential loop body a given 
number of times. In this case, both the loop body and the loop itself are 
{\tt Thread} Programs. More interesting is the {\tt ForAll} case that introduces 
 different kinds of parallelism depending on the {\tt t} parameter.
{\tt ForAll} executes a program body at the level below several times in parallel. 
If the body of the {\tt ForAll} loop is a {\tt Thread} computation the {\tt ForAll} loop 
itself is a {\tt Block} computations. Likewise, if the body is a {\tt Block} computation, 
the loop itself is a {\tt Grid} computation. 
%% {\tt ForAllBlocks} executes several blocks 
%% in parallel over a {\tt Grid}. The {\tt Thread}, {\tt Grid} and {\tt Block} 
%% parameters rule out any nesting of loops that we cannot easily compile 
%% to GPU code.

@@

A push array has two parts, a length and a higher-order function. 
The idea is that this function is provided with a method to write an 
element into memory (a {\em write-function}), and returns a program 
that uses that write-function a number of times. Note that the 
write-function produces a {\tt TProgram} (short for {\tt Program Thread}), This 
restricts our push arrays from representing nested data parallelism. 
This concept is best illustrated by showing how to turn a pull array into a push array.

@@

The {\tt ASize} class has instances for both the static and dynamic lengths 
and is used to obtain conversion to a representation used internally; it 
provides a function called {\tt sizeConv} which can convert both static and 
dynamic lengths to a {\tt Word32} expression.

@@

The last step in compilation process uses the Language.C.Quote library to 
turn the {\tt IM} into CUDA code. This compilation stage takes a {\em configuration} 
as parameter. The configuration tells how many actual threads per block the 
CUDA code should be generated for. At this stage parallel for loops ({\tt SForAll})
are turned into more than one parallel for loop in sequence, if needed.

@@

\begin{figure} 
\begin{small}
\centering
\begin{tabular}{| l | c | c | c | c | r | }
\hline 
  Kernel & Threads & Shared mem. & ms \\ \hline 
  red1 & 1024 & 16384 & 2.085 \\ \hline
  red2 & 1024 & 16384 & 1.748 \\ \hline
  red3 & 1024 & 16384 & 1.741 \\ \hline
  red4 & 256 & 2048  & 2.113  \\ \hline
  red5 & 256 & 2048  & 0.455  \\ \hline
  red6 & 128 & 1024  & 0.441  \\ \hline
  red7 & 64 & 512    & 0.441  \\ \hline
\end{tabular}
\caption{ Execution time of a grid of 8192 reductions , each of
2048 elements on 
and on a GTX680. The timing figures were obtained using the NVIDIA Profiler.}
\label{fig:reducetable} 
\end{small} 
\end{figure}

@@

%% \end{tabular}
%% \caption{CAPTION CAPTION CAPTION}
%% \label{fig:scantable} 
%% \end{small} 
%% \end{figure} 
\begin{figure}
\begin{minipage}{.5\linewidth}
\begin{tikzpicture} [scale = 0.5]
\begin{axis} [xmax = 3, 
              xmin = 0, 
              ymax = 6,
              ymin = 0,
              % ytick = data,
              xtick = {0,1,2,3},
              xticklabels = {128,256,512,1024},
              legend entries = {Sklansky, Sklansky2, Sklansky3},
              legend pos = north west,
              xlabel = \#Elements, 
              ylabel = Time in ms,
              title = 4096 Blocks,
  ]         
  \addplot[color = black, 
           mark  = *,
           mark size = 2pt] coordinates {
    (0, 0.364 )
    (1, 0.847 )
    (2, 2.098 ) 
    (3, 5.762 )
  };
  \addplot[color = black, 
           mark  = square*,
           mark size = 2pt] coordinates {
    (0, 0.408 )
    (1, 0.564 )
    (2, 1.292 ) 
    (3, 3.336 )
  };
  \addplot[color = black, 
           mark  = diamond*,
           mark size = 2pt] coordinates {
    (0, 0.341 )
    (1, 0.514 )
    (2, 1.179 ) 
    (3, 2.957 )
  };

\end{axis}
\end{tikzpicture} 
\end{minipage}
\begin{minipage}{.5\linewidth}
\begin{tikzpicture} [scale = 0.5]
\begin{axis} [xmax = 3, 
              xmin = 0, 
              ymax = 6,
              ymin = 0,
              % ytick = data,
              xtick = {0,1,2,3},
              xticklabels = {128,256,512,1024},
              legend entries = {Sklansky, Sklansky2, Sklansky3},
              legend pos = north west,
              xlabel = \#Elements,
              title = N/BlockSize Blocks
  ]         
  \addplot[color = black, 
           mark  = *,
           mark size = 2pt] coordinates {
    (0, 2.882 )
    (1, 3.372 )
    (2, 4.194 ) 
    (3, 5.762 )
  };
  \addplot[color = black, 
           mark  = square*,
           mark size = 2pt] coordinates {
    (0, 3.226 )
    (1, 2.242 )
    (2, 2.576 ) 
    (3, 3.336 )
  };
  \addplot[color = black, 
           mark  = diamond*,
           mark size = 2pt] coordinates {
    (0, 2.723 )
    (1, 2.041 )
    (2, 2.352 ) 
    (3, 2.957 )
  };

@@

\end{minipage}
\caption{\emph{left:}The time it takes to run 4096 blocks of varying sizes using our
different implementations of the Sklansky kernel. \newline
\emph{Right:} varies the number of blocks so that the number of elements is the 
same in each experiment. These results where obtained on an NVIDIA GT 650m using the 
NVIDIA Profiler.}
\label{fig:scangraphs}
\end{figure}

@@

\subsubsection{To automate or not to automate} 
Obsidian makes use of types that model the GPU hierarchy. This limits 
the kind of programs one can express but also ensures that what we can write 
is suitable for the GPU. An alternative is to remove these types and at the 
same time make the compiler more clever. This would move some of the 
required GPU detail knowledge from the programmer into compiler transformations. 
We have recently started a project to investigate this change of Obsidian
together with a Master's student.

\subsubsection{To optimise or not to optimise} 
The version of Obsidian described here does not try to use any compiler optimisations techniques. Instead, we are expecting that the 
CUDA compiler will apply a good set of techniques, from common subexpression 
elimination to more GPU specific transformations. As future work, we 
will investigate to what extent this reliance on the CUDA
compiler is justified. Also, we want to 
see if there are high-level optimisations that we can perform more easily 
on our intermediate representations of GPU kernels than what is possible 
on low-level CUDA code.

@@

In section \ref{sec:Benchmarks}, we implemented large reduction using 
CUDA and kernels generated using Obsidian. This is something we want to 
be able to avoid. We envision a kernel coordination language embedded 
in Haskell where the programmer can express the kind of sequential 
and parallel compositions of kernels; note that this is different from 
composing two local computations into one, which we already have the 
capability to do. Having an embedded language for expressing full 
algorithms that make use of many different kernels would further 
improve the usability of Obsidian. We have started a project with the 
goal of implementing this language together with a Master's student.

>>>>

The NVIDIA CUDA capable GPUs are built on a scalable architecture. A GPU 
consists of a number of {\em multiprocessors}; each multiprocessor has a 
number of processing elements (cores) and a local memory that is shared 
between threads running on the cores. Figure \ref{fig:GPU} illustrates these 
concepts. A GPU can come with as few as one of these multiprocessors.
The GPUs used in our measurements are the GTX 680, which has eight 
multiprocessors, with a total of 1536 processing cores, the GTX 670 with 1344 cores, 
and the GT 650M, a laptop GPU, which has 384 such cores. On these cores, 
groups of 32 threads called {\em warps} are scheduled. There are a number 
of warp scheduling units per multiprocessor. Within a warp, threads execute 
in lockstep (SIMD); diverging branches, that is those that take different 
paths on different threads within a warp, are serialised, leading to 
performance penalties.

@@

Obsidian, like CUDA, differentiates between {\em Thread}, {\em Block} and 
{\em Grid} computations. The programmer specifies a sequential computation 
to be carried out by each thread. Many instances of a sequential computation  
run in parallel over threads, to form a block. A number of such blocks run in 
parallel, to form a grid.

@@

This program does not make use of any specific details of the BProgram monad; 
instead it just composes {\tt return} and {\tt push} with the Obsidian library 
function {\tt reverse}. The {\tt push} function converts a pull array to a push 
array. {\tt SPull} and {\tt SPush} arrays have static lengths, this is explained in more 
detail on page \pageref{pag:HLSTATIC}.

@@

\pagebreak
\noindent
The {\tt pConcatMap} combinator is overloaded; here it is applied 
at the following type: 
%takes a local computation that results in a push array.
\begin{small}
\begin{Verbatim}[samepage=true] 
pConcatMap :: ASize l 
            => (a1 -> Program t (SPush t a))
            -> Pull l a1 
            -> Push (Step t) l a
\end{Verbatim}
\end{small}

@@

\noindent
where {\tt a0} to {\tt aN} can be pull arrays or scalars. This choice 
limits the kinds of kernels can be expressed, but
makes compilation of those kernels a lot easier.
For example, a kernel that outputs data to two different global
arrays cannot currently be expressed. Lifting this restriction 
is work in progress.

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\citehl{COMPILEEDSL} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\citehl{FELDSPAR} 
and Repa ~\citehl{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as needed from those available. 
If the array is larger than the number of threads made available, the 
force operation is split into two or more passes. This is a change compared
to earlier versions of Obsidian, where the length of arrays we could force 
were limited to the maximum number of threads per block (1024 threads). In 
this version, the threads are, in a sense, virtual. The mapping from virtual 
threads to real threads is done as a step in the compilation.

@@

There are type synonyms for the dynamic and static versions of  pull arrays: \label{pag:HLSTATIC}

@@

The programs that can be specified in Obsidian are represented in a {\tt Program} data type.
This data type also represents the lowest level programming interface 
exposed by Obsidian. In reference ~\citehl{JOSEFJOELCSORT} we make use of the low-level 
capabilities provided by that interface to implement histograms using atomic 
operations (not shown here).

@@

Below are some selected parts of the  {\tt Program} data type. There 
is a sequential, {\tt SeqFor}, loop that runs a sequential loop body a given 
number of times. In this case, both the loop body and the loop itself are 
{\tt Thread} Programs. More interesting is the {\tt ForAll} case that introduces 
 different kinds of parallelism depending on the {\tt t} parameter.
{\tt ForAll} executes a program body at the level below several times in parallel. 
If the body of the {\tt ForAll} loop is a {\tt Thread} computation, the {\tt ForAll} loop 
itself is a {\tt Block} computations. Likewise, if the body is a {\tt Block} computation, 
the loop itself is a {\tt Grid} computation. 
%% {\tt ForAllBlocks} executes several blocks 
%% in parallel over a {\tt Grid}. The {\tt Thread}, {\tt Grid} and {\tt Block} 
%% parameters rule out any nesting of loops that we cannot easily compile 
%% to GPU code.

@@

A push array has two parts, a length and a higher-order function. 
The idea is that this function is provided with a method to write an 
element into memory (a {\em write-function}), and returns a program 
that uses that write-function a number of times. Note that the 
write-function produces a {\tt TProgram} (short for {\tt Program Thread}). 
As a result, push arrays cannot represent nested data parallelism. 
This concept is best illustrated by showing how to turn a pull array into a push array.

@@

The {\tt ASize} class has instances for both the static and dynamic lengths 
and is used to obtain conversion to a representation used internally; 
the function {\tt sizeConv} converts both static and dynamic lengths to a {\tt Word32} expression.

@@

The last step in the compilation process uses the Language.C.Quote library to 
turn the {\tt IM} into CUDA code. This compilation stage takes a {\em configuration} 
as parameter. The configuration tells how many actual threads per block the 
CUDA code should be generated for. At this stage parallel for loops ({\tt SForAll})
are turned into more than one parallel for loop in sequence, if needed.

@@

\begin{figure} 
\begin{small}
\centering
\begin{tabular}{| l | c | c | c | c | r | }
\hline 
  Kernel & Threads & Shared mem. & ms \\ \hline 
  red1 & 1024 & 16384 & 2.085 \\ \hline
  red2 & 1024 & 16384 & 1.748 \\ \hline
  red3 & 1024 & 16384 & 1.741 \\ \hline
  red4 & 256 & 2048  & 2.113  \\ \hline
  red5 & 256 & 2048  & 0.455  \\ \hline
  red6 & 128 & 1024  & 0.441  \\ \hline
  red7 & 64 & 512    & 0.441  \\ \hline
\end{tabular}
\caption{ Execution time of a grid of 8192 reductions , each of
2048 elements and on a GTX680. The timing figures were obtained using the NVIDIA Profiler.}
\label{fig:reducetable} 
\end{small} 
\end{figure}

@@

%% \end{tabular}
%% \caption{CAPTION CAPTION CAPTION}
%% \label{fig:scantable} 
%% \end{small} 
%% \end{figure} 
\begin{figure}
\begin{minipage}{.5\linewidth}
\begin{tikzpicture} [scale = 0.6]
\begin{axis} [xmax = 3, 
              xmin = 0, 
              ymax = 1.5,
              ymin = 0,
              % ytick = data,
              xtick = {0,1,2,3},
              xticklabels = {128,256,512,1024},
              legend entries = {Sklansky, Sklansky2, Sklansky3},
              legend pos = north west,
              xlabel = \#Elements, 
              ylabel = Time in ms,
              title = 4096 Blocks,
  ]         
  \addplot[color = black, 
           mark  = *,
           mark size = 2pt] coordinates {
    (0, 0.096 )
    (1, 0.214 )
    (2, 0.517 ) 
    (3, 1.382 )
  };
  \addplot[color = black, 
           mark  = square*,
           mark size = 2pt] coordinates {
    (0, 0.076 )
    (1, 0.126 )
    (2, 0.286 ) 
    (3, 0.714 )
  };
  \addplot[color = black, 
           mark  = diamond*,
           mark size = 2pt] coordinates {
    (0, 0.077 )
    (1, 0.127 )
    (2, 0.286 ) 
    (3, 0.694 )
  };

\end{axis}
\end{tikzpicture} 
\end{minipage}
\begin{minipage}{.5\linewidth}
\begin{tikzpicture} [scale = 0.6]
\begin{axis} [xmax = 3, 
              xmin = 0, 
              ymax = 1.5,
              ymin = 0,
              % ytick = data,
              xtick = {0,1,2,3},
              xticklabels = {128,256,512,1024},
              legend entries = {Sklansky, Sklansky2, Sklansky3},
              legend pos = north west,
              xlabel = \#Elements,
              title = N/BlockSize Blocks
  ]         
  \addplot[color = black, 
           mark  = *,
           mark size = 2pt] coordinates {
    (0, 0.715 )
    (1, 0.832 )
    (2, 1.025 ) 
    (3, 1.381 )
  };
  \addplot[color = black, 
           mark  = square*,
           mark size = 2pt] coordinates {
    (0, 0.551 )
    (1, 0.475 )
    (2, 0.562 ) 
    (3, 0.715 )
  };
  \addplot[color = black, 
           mark  = diamond*,
           mark size = 2pt] coordinates {
    (0, 0.560 )
    (1, 0.487 )
    (2, 0.565 ) 
    (3, 0.692 )
  };

@@

\end{minipage}
\caption{\emph{left:}The time it takes to run 4096 blocks of varying sizes using our
different implementations of the Sklansky kernel. \newline
\emph{Right:} varies the number of blocks so that the number of elements is the 
same in each experiment. These results where obtained on an NVIDIA GTX 670.}
\label{fig:scangraphs}
\end{figure}

@@

\subsubsection{To automate or not to automate} 
Obsidian makes use of types that model the GPU hierarchy. This limits 
the kind of programs that can be expressed, ensuring that the resulting programs are
suitable for the GPU. An alternative is to remove these types and at the 
same time make the compiler more clever. This would move some of the 
required GPU detail knowledge from the programmer into compiler transformations. 
We have recently started a project to investigate this change to Obsidian,
together with a Masters student.

\subsubsection{To optimise or not to optimise} 
The version of Obsidian described here does not try to use any compiler 
optimisation techniques. Instead, we are expecting that the 
CUDA compiler will apply a good set of techniques, from common subexpression 
elimination to more GPU specific transformations. As future work, we 
will investigate to what extent this reliance on the CUDA
compiler is justified. Also, we want to 
see if there are high-level optimisations that we can perform more easily 
on our intermediate representations of GPU kernels than what is possible 
on low-level CUDA code.

@@

In section \ref{sec:Benchmarks}, we implemented large reduction using 
CUDA and kernels generated using Obsidian. This is something we want to 
be able to avoid. We envision a kernel coordination language embedded 
in Haskell where the programmer can express the kind of sequential 
and parallel compositions of kernels; note that this is different from 
composing two local computations into one, which we already have the 
capability to do. Having an embedded language for expressing full 
algorithms that make use of many different kernels would further 
improve the usability of Obsidian. We have started a project with the 
goal of implementing this language, together with a Masters student.

### 2013-11-12 src2/thesis.tex
update

We have implemented basic algorithms such as sorting, merging, fractal generation, reductions 
and parallel prefix (scans). Often we have managed to get very good performance, close to that 
of hand optimised CUDA code. However, there are also cases where we just cannot seem to reach 
the level of performance that we would want. One such example is parallel prefix, an 
application that has been haunting us from the very beginning. Scan is problematic to us because 
it is best implemented as an in-place algorithm, while our array representations do 
not allow for in-place update; there is no memory associated with them to update. 
We are trying to improve this situation by adding a third array representation to Obsidian,
mutable arrays.

Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
a reduction kernel in Obsidian. At the end, the implemented reduction kernel is  
faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
been happy just getting close.

@@

\FloatBarrier
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Future work
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Future work}

>>>>

We have implemented basic algorithms such as sorting, merging, fractal generation, reductions 
and parallel prefix (scans). Often we have managed to get very good performance, close to that 
of hand optimised CUDA code. GPU execution of scan operations is something that has 
been worked on extensively in the GPGPU community and very efficient implementations 
exist~\citet{merrill,ScanCUDA,OLAMARCUS,EFFICIENTSCAN}. In Paper F (section~\ref{sec:paperF}), 
we only study scan kernels (scan of small arrays) and measure their speed relative to eachother. 
Compared to a scan kernel supplied with the CUDA API, our kernel fastest kernel is slighty more 
than three times slower. Performance needs to be improved at the kernel level before we attempt 
composing them into large scan operations. We hope to improve this situation with the 
addition of the warp level abstractions in Obsidian (this is future work, 
section~\ref{sec:thesisFUTURE}).

Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
a reduction kernel in Obsidian. At the end, the implemented reduction kernel 
reaches a throughput of 140GB/sec on a GPU with 192GB/sec memory bandwith. 
%faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
%reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
%been happy just getting close.

@@

\FloatBarrier
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Future work
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Future work} 
\label{sec:thesisFUTURE}
